Supplement History of Cheminformatics as part of Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-08 supplement_history.tex
more updates to the suppl

\title{Supplement History of Chem(o)informatics as part of Cheminformatics: The Computer Science of Chemical Discovery}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockville, MD 20850}\\
\email{guhar@mail.nih.gov}
}

@@

\section{History of Chem(o)informatics}
The aim of this brief historic review of cheminformatics is to put the
content of the main article into a broader perspective. Therefore it
might be useful to clarify some other terms and how cheminformatics
might relate to them.  Still, Chen \cite{Chen2006} and Brown
\cite{brown2009} cited various definitions of pioniers in the field
and Brown concluded rightfully: '\textit{The differences in
  definitions (of the term cheminformatics) are largely a result of
  the types of analyses that particular scientists practice and no
  single definition is intended to be all-encompassing}.'
\begin{itemize}
\item \textbf{Quantum chemistry} is something most chemists would
  relate with 'theoretical chemistry'. In short, it allows to describe
  chemical molecules on an electron density level, which allows also
  to describe physical properties, e.g infrared spectra of a chemical
  compound.  The results of quantum chemistry calculation are used in
  molecular modeling (energy force field parameters) and
  cheminformatics (atom partial charges, or full molecule properties,
  e.g molecule polarizabilities).
\item \textbf{Molecular modeling} or \textbf{Computational chemistry}
  considers typically energy calculations of single molecules or
  interactions with other molecules, e.g. proteins. 'Molecular
  docking' belongs into this category, since it places a chemical
  molecule in the active site of enzymes and calculates
  energy/interaction scores between a chemical molecule and the
  protein. Surely are there various cheminformatics concepts required
  to perform such a task. Still, strictly speaking is cheminformatics
  rather on the chemical molecule side than on the protein side,
  though are both sides affecting each other. This area is typically
  strongly related to structural biology and technologies like XRay
  crystallography and NMR spectra structure determination.
\item \textbf{Chemometrics} can be considered as applied
  cheminformatics to the field of chemical/physical analysis,
  e.g. infrared, mass and NMR spectra analysis.
\item '\textbf{Translational research} aims to improve the health and
  longevity of the world's populations and depends on developing
  broad-based teams of scientists and scholars who are able to focus
  their efforts to link basic scientific discoveries with the arena of
  clinical investigation, and translating the results of clinical
  trials into changes in clinical practice, informed by evidence from
  the social and political sciences.' (Source:Wikipedia).  Since
  clinical trials can be connected to drugs, chemical molecules, even
  all of translational research and biobanking could be considered as
  a field of cheminformatics, which would be on the other hand just a
  small contribution to a full knowledge management in this field
  \cite{Szalma2010}.
\end{itemize}
If we consider all information, analysis, and optimization of a
chemical molecule as 'cheminformatics', then the field is very large,
since a chemical molecule plays the central role in many related
disciplines. Still, for keeping it practical the aim of this review is
not being exhaustive, but rather to point the reader to more detailed
references and to highlight some historic milestones.

>>>>

\title{Supplement History of Cheminformatics as part of Cheminformatics: The Computer Science of Chemical Discovery}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockville, MD 20850}\\
\email{guhar@mail.nih.gov}
}

@@

\section{History of Cheminformatics}
The aim of this brief review of the history of cheminformatics is to
put the content of the main article into a broader perspective. In
addition, we also discuss some terms as they relate to
cheminformatics.  It is thus useful to first provide some definitions
of cheminformatics itself, given that they also shed some light on the
founding principles of the field. For example, we have the following
definitions:
\begin{itemize}
\item ``\textit{The mixing of information resources to transform data
    into information, and information into knowledge, for the intended
    purpose of making better decisions faster in the arena of drug
    lead identification and optimization.}'' Frank K. Brown, 1998.
\item ``\textit{[Chemoinformatics involves]... the computer
    manipulation of two- or threedimensional chemical structures and
    excludes textual information. This distinguishes the term from
    chemical information, largely a discipline of chemical librarians
    and does not include the development of computational methods.}''
  Peter Willett, 2002.
\item ``\textit{\ldots the application of informatics to solve chemical
    problems.}'' and ``\textit{\ldots chemoinformatics makes the point that
    you're using one scientific discipline to understand another
    scientific discipline.}'' Johann Gasteiger, 2002
\end{itemize}
More generally, Chen\cite{Chen2006} and Brown \cite{brown2009} cited various
definitions of pioneers in the field with Brown concluding that
``'\textit{The differences in definitions (of the term
  cheminformatics) are largely a result of the types of analyses that
  particular scientists practice and no single definition is intended
  to be all-encompassing}'', as is evident from the three definitions above.

It is also useful to clarify some other terms and how
cheminformatics might relate to them:
\begin{itemize}
\item \textbf{Quantum chemistry} is generally associated with
  theoretical chemistry or chemical physics. Briefly, it focuses on
  the description of chemical systems starting from irst principles
  and is based on the Schr\"{o}dinger wave equation and frameworks
built on top of it. The use of quantum mechanics allows one to
describe molecules in terms of electron density, which thus forms the
basis of physically accurate calculations of physical properties
(e.g., infrred spectra). The results of quantum chemistry calculations
are used in molecular modeling (energy force field parameters) and
cheminformatics (atom partial charges, or full molecule properties,
e.g molecule polarizabilities).
\item \textbf{Molecular modeling} or \textbf{Computational chemistry} considers typically energy calculations of single molecules or interactions
with other molecules, e.g. proteins. 'Molecular docking' belongs into this category, since it places a chemical molecule in the active site of enzymes
and calculates energy/interaction scores between a chemical molecule and the protein. Surely are there various cheminformatics concepts required to
perform such a task. Still, strictly speaking is cheminformatics rather on the chemical molecule side than on the protein side, though are both sides
affecting each other. This area is typically strongly related to structural biology and technologies like XRay crystallography and 
NMR spectra structure determination. 
\item \textbf{Chemometrics} can be considered as applied cheminformatics to the field of chemical/physical analysis, e.g. infrared, mass and NMR spectra analysis.
\item '\textbf{Translational research/medicine} aims to improve the health and longevity of the world's populations and depends on developing broad-based 
teams of scientists and scholars who are able to focus their efforts to link basic scientific discoveries with the arena of clinical investigation, 
and translating the results of clinical trials into changes in clinical practice, informed by evidence from the social and political sciences.' (Source:Wikipedia).
Since clinical trials can be connected to drugs, chemical molecules, even all of translational research and biobanking could be considered as a field
of cheminformatics, which would be on the other hand just a small contribution to a full knowledge management in this field \cite{Szalma2010}.
\end{itemize}
If we consider all information, analysis, and optimization of a chemical molecule as 'cheminformatics', then the field is very large, since a chemical
molecule plays the central role in many related disciplines. Still, for keeping it practical the aim of this review is not being exhaustive, but rather to 
point the reader to more detailed references and to highlight some historic milestones.
